\documentclass[runningheads,envcountsame]{llncs}
\usepackage[T1]{fontenc}
\usepackage{url}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{mathtools}
\usepackage[dvipsnames]{xcolor}
\usepackage{wasysym}
\usepackage{annotate-equations}
\usepackage{quiver}
\usepackage{cleveref}
\usepackage{ebproof}
\usepackage[T1]{fontenc}
\usepackage{wasysym}
\usepackage{multirow}

% comments with
% https://mirror.ox.ac.uk/sites/ctan.org/macros/latex/contrib/todonotes/todonotes.pdf
\usepackage[colorinlistoftodos]{todonotes}


%%%% configuration and new commands  %%%%
\input{commands}
\input{listing-config}

\renewcommand{\floatpagefraction}{.99}

\begin{document}

% \title{Verifying SMT-Based Linear Integer Arithmetic in Lambdapi}
% \title{SMT Linear Integer Arithmetic proof checking in Lambdapi}
\title{Checking Linear Integer Arithmetic Proofs\\ in Lambdapi}

% \titlerunning{Verification of LIA proof with Lambdapi}

\author{Alessio Coltellacci\inst{1}\orcidID{0009-0005-3580-2075}
  \and
  Stephan Merz\inst{1}\orcidID{0000-0003-0974-1844}
}
%
\authorrunning{A. Coltellacci, S. Merz}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of Lorraine, CNRS, Inria, LORIA, Nancy 
%\email{alessio.coltellacci@inria.fr}
}

% \institute{University of Lorraine, CNRS, Inria, LORIA \and
% Nancy
% \email{alessio.coltellacci@inria.fr}\\
% \url{http://www.springer.com/gp/computer-science/lncs} \and
% ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
% \email{\{abc,lncs\}@uni-heidelberg.de}}
%
\maketitle
%
\begin{abstract}
Modern SMT solvers can generate proofs of unsatisfiability so that the result can be checked independently.
A dependable approach to verify these proofs is to reconstruct them within a proof assistant.
In previous work, the SMT checker Carcara was extended to reconstruct SMT proofs in Lambdapi — a proof assistant designed for interoperability,
supporting the import and export of proofs for integration with other proof assistants such as Rocq, Lean, or HOL-Light.
Whereas that work was limited to SMT theories without arithmetic, we here present an extension that enables the reconstruction of SMT proofs involving linear integer arithmetic.

\keywords{SMT \and Alethe \and integer arithmetic \and Lambdapi \and normal form \and proof by reflection}
\end{abstract}

\section{Introduction}

SMT solvers have become capable of producing proofs of unsatisfiability,  enabling independent verification of correctness without relying on the solver's implementation.
This development addresses the growing need for trustworthy verification in safety-critical applications and for using SMT solvers as backends of skeptical proof assistants that do not trust the solver.
Alethe \cite{alethe,alethespec} is an established SMT proof format supported by the solvers cvc5 and veriT. In our previous work \cite{ColtellacciMD24}, we extended the Alethe proof checker and elaborator Carcara \cite{carcara} to
reconstruct Alethe SMT proofs within the Lambdapi proof assistant, thereby ensuring their validity.
Lambdapi \cite{lambdapi} is a proof assistant based on the $\lambda\Pi$-calculus modulo rewriting \cite{lpmodulo}, a logical framework \cite{lf} that extends the $\lambda$-calculus with dependent types and user-defined rewrite rules.
This foundation allows Lambdapi to serve as a framework for formalizing various logical systems.
Designed with interoperability in mind, Lambdapi can import and export proofs, facilitating integration with other proof assistants such as Rocq \cite{Rocq-refman}, Lean \cite{lean4:2021}, and HOL-Light \cite{LPAR2024:Translating_HOL_Light_proofs}. 

However, our prior work was limited to proofs expressed within the logic of Uninterpreted Functions (UF).
In the present work, we extend this approach to support the reconstruction of proof steps involving linear integer arithmetic (LIA).
In the context of LIA, the SMT solver determines unsatisfiability by analyzing formulas composed of integer variables, linear arithmetic operations, and logical connectives.
This process typically relies on techniques such as Simplex methods adapted for integers and cutting-plane methods to establish the unsatisfiability of linear constraints \cite{SRI:simplex:dpllt,bromberger:lia}.
Our main contribution is the implementation of an automatic decision procedure to verify these proof steps, using proof by reflection \cite{reflection-origin-coq,ring-coq}.

The remainder of this paper is organized as follows.
In Sect.~\ref{sec:background}, we provide an overview of Lambdapi and present the structure of Alethe proof certificates for linear arithmetic.
Then, in Sect.~\ref{sec:elaboration-lia}, we describe how we leverage Carcara's elaboration process to reconstruct linear integer arithmetic steps, even when coefficient annotations are missing.
Section~\ref{sec:encoding} introduces our encoding of SMT linear arithmetic within Lambdapi, while Sect.~\ref{sec:lia-reconstruction} presents an automatic decision procedure, based on proof by reflection, to verify these arithmetic steps.
An empirical evaluation of our approach is provided in Sect.~\ref{sec:evaluation}. We review related work in Sect.~\ref{sec:related}, and we conclude in Sect.~\ref{sec:conclusion}.

\input{background}

\input{elaboration}

\input{lambdapi}

\input{lia-reconstruction}

\input{evaluation}

\section{Related work}
\label{sec:related}

The reconstruction for Alethe proofs produced by the veriT solver has been implemented within Isabelle/HOL \cite{aletheInIsa} to enhance the reconstruction success rate compared to Z3.
% sm: No – the paper clearly says that linarith was previously used for reconstructing Z3 proofs but that its scope was limited.
%The authors employed Isabelle's \emph{linarith} procedure to replay arithmetic reasoning steps. This tactic serves as a decision procedure over real numbers, rather than integers or natural numbers, and is based on Fourier-Motzkin elimination.
%However, the authors note that Isabelle's arithmetic procedure is incomplete and relatively inefficient for reconstructing arithmetic proofs; consequently, specific rules are reconstructed heuristically to improve performance.
Concerning linear arithmetic reasoning, they also follow the algorithm described in Sect.~\ref{ssect:alethe}, relying on Isabelle's simplifier (which supports reasoning modulo associativity and commutativity) for normalizing expressions.

Besson \cite{micromega} introduces a modular framework for integrating SMT solvers with the Rocq proof assistant, emphasizing the generation and efficient verification of proof certificates directly within Coq.
His approach enables fast, reflexive proof checking using Nelson-Oppen-style theory combination, and it supports both Linear Integer Arithmetic (LIA) and Linear Arithmetic (LA)\todo[sm]{Not clear what that is -- LRA? Also, it appears that this is necessary because they don't get the parameters from the SMT proof.} logics.
To efficiently generate Farkas certificates for LA, the technique relies on the Simplex algorithm, while for LIA, a variant of the Omega test \cite{omegatest} is adopted.
The technique has been evaluated on SMT-LIB 2 benchmarks using the Z3 solver.

Compared to these approaches, which demonstrate stronger scalability in handling larger and more complex systems of inequalities, our method does not yet scale as well, in particular for cases that require significant arithmetic simplification.
As shown in Sect.~\ref{sec:evaluation}, performance degrades on benchmarks such as \emph{rings}, where large instances of rule \tt{la\_generic} are generated.
Nevertheless, our framework offers the advantage of supporting proof export from Lambdapi to other systems. This interoperability enables the reuse of valid SMT proofs in external environments such as Rocq.


\section{Conclusion}
\label{sec:conclusion}

We presented an extension of Carcara's \cite{carcara}\footnote{\url{https://github.com/NotBad4U/carcara/tree/lambdapi-translate}}
 translation module \cite{ColtellacciMD24} for reconstructing arithmetic reasoning steps in Alethe proof traces involving linear arithmetic reasoning in the foundational proof assistant Lambdapi \cite{lambdapi}.
The translation module enables the export of an Alethe proof, together with its input problem, parsed and elaborated by Carcara, into a Lambdapi file that can be independently verified using the Lambdapi compiler and the Lambdapi library included in Carcara, which encodes Alethe logic.
Building on this module, our extension introduces support for exporting arithmetic proof steps and extends the Lambdapi Alethe library to incorporate integer arithmetic proof rules.
Our method includes a ring solver inspired by the computational reflection techniques described in \cite{reflection-origin-coq,ring-coq}.

We have shown that this allows us to reconstruct several thousand SMT proofs that originate from the SMT-lib Benchmarks.
To the best of our knowledge, this is the first work to automatically solve arithmetic goals in Lambdapi.
Nonetheless, the current implementation does not yet scale well in arithmetic-heavy cases.
As future work, we plan to improve Lambdapi's built-in normalization mechanism to better handle such cases.
We also intend to extend our work to support rational and real arithmetic, enabling verification of broader fragments of linear arithmetic (LRA) theories.
This will in particular require an encoding of rational and real arithmetic in Lambdapi and their normalization function, based on the $\mathit{gcd}$ operator, as a basis for implementing proof by reflection.

\bibliographystyle{splncs04}
\bibliography{refs}

%\include{appendix}

\end{document}
